---
title: "Project 1 Obesity Prevalence"
author: "Shihong Wang, Jamie Watt, Mackenzie Whitelock, Irene Hanna Anu"
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(sjPlot)
library(kableExtra)

```

# Introduction {#sec-intro}

Obesity is a significant public health concern that affects morbidity and mortality. Monitoring its prevalence is essential for evaluating public health policies and interventions. The Scottish Health Survey, which samples individuals from private households across Scotland, provides an opportunity to study trends in obesity. The primary goal of this analysis is to determine whether there has been a statistically noticeable change in obesity prevalence in Scotland over the period 2008–2012. The secondary goal of this analysis is to assess whether there is a difference in obesity levels in other variables recorded in the Scottish Health Survey between 2008 and 2012.

# Exploratory Analysis {#sec-EA}


## Prevalance of Obesity Through The Years


The prevalence of obesity for each year was computed using the formula:

$$
\text{Prevalence (in%)} = \frac{N_{\text{obese}}}{N_{\text{total}}} \times 100
$$

First we will produce a line plot with data points to illustrate the trend in obesity prevalence over the years.

```{r}
data <- read.csv("DAProject1.csv")

data$Year <- as.factor(data$Year)

prevalence <- data %>%
  group_by(Year) %>%
  summarise(total = n(),
            obese_count = sum(Obese == "Yes"),
            prevalence = (obese_count / total) * 100)

ggplot(prevalence, aes(x = Year, y = prevalence)) +
  geom_line(group = 1, color = "black") +
  geom_point(size = 3, color = "red") +
  labs(title = "Obesity Prevalence Over Time in Scotland",
       x = "Year",
       y = "Prevalence (in %)")
```

From the graph we can see that the prevalence appears to fluctuate year to year, rather than moving in a strictly upward or downward path. This suggests that obesity rates may have been influenced by a range of factors that vary from year to year (e.g., sample composition, lifestyle changes, policy impacts).

Also we will produce a bar plot to display the count of individuals classified as obese versus not obese for each survey year.

```{r}
ggplot(data, aes(x = Year, fill = Obese)) +
  geom_bar(position = "dodge") +
  labs(title = "Count of Obesity Classification by Year",
       x = "Year",
       y = "Count") 
```

We can observe that in each year, the “No” bar (non-obese) is taller than the “Yes” bar (obese). However, the gap between these two bars may differ slightly from year to year.

## Differences in Obesity Through The Years

```{r}
plot.edu <- ggplot(data, aes(x = Education, fill = Obese)) +
  geom_bar(position = "fill") +
  labs(title = "Obesity by Education Level", x = "Education Level", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values=c("dodgerblue4", "firebrick3"))

plot.veg <- ggplot(data, aes(x = Veg, fill = Obese)) +
  geom_bar(position = "fill") +
  labs(title = "Obesity and Vegetable Intake", x = "Vegetable Intake", y = "Proportion")+
  scale_fill_manual(values=c("dodgerblue4", "firebrick3"))

plot.fruit <- ggplot(data, aes(x = Fruit, fill = Obese)) +
  geom_bar(position = "fill") +
  labs(title = "Obesity and Fruit Intake", x = "Fruit Intake", y = "Proportion")+
  scale_fill_manual(values=c("dodgerblue4", "firebrick3"))

plot.age1 <- ggplot(data, aes(y = (Age), x = Obese, fill = Obese)) +
  scale_fill_manual(values = c("dodgerblue4", "firebrick3")) +
  geom_boxplot()

plot.age2 <- ggplot(data, aes(x = (Age), fill = Obese, alpha = 0.1)) +
  scale_fill_manual(values = c("dodgerblue4", "firebrick3")) +
  geom_density()

plot.year <- ggplot(data, aes(x=Year, fill=Obese))+
  geom_bar(position="fill")+
  scale_fill_manual(values = c("dodgerblue4", "firebrick3"))+
  labs(y="Proportion")

plot.sex <- ggplot(data, aes(x=Sex, fill=Obese))+
  geom_bar(position="fill")+
  scale_fill_manual(values = c("dodgerblue4", "firebrick3"))+
  labs(y="Proportion")
```

We want to see whether initial plots may give an indication to if there is in fact a difference in obesity levels across the other variables in the survey. We will begin by looking at fruit and veg intake to see if there are differences in these areas.

```{r}
#| label: exp plots 1
#| fig-cap: Proportion of Obesity by Fruit and Veg intake

gridExtra::grid.arrange(plot.fruit, plot.veg, ncol = 2)
```

We can see here that the proportions of obesity appear to be almost in the same in fruit in take, although there may be a slight difference in whether a person consumes vegetables.

Next lets look at differences across years and sex.

```{r}
#| label: exp plots 2
#| fig-cap: Proportion of Obesity by Year and Sex

gridExtra::grid.arrange(plot.year, plot.sex, ncol = 2)
```

Now we can see that the proportion of obesity across years 2008-2012 do fluctuate slightly but not a lot so we will need to look at this further. There appears to be almost no difference in levels of obesity across males and females though, as these proportions are almost identical.

Next we shall look at whether there are differences in obesity levels across ages.

```{r}
#| label: exp plots 3
#| fig-cap: Boxplot and Density plot of age

gridExtra::grid.arrange(plot.age1, plot.age2, ncol = 2)
```

We can see here that the age of obese people tends to be slightly older than those who are not obese, however we cannot tell whether this is a significant difference from these plots. 

Finally, we will look at obesity levels across different levels of education.

```{r}
#| label: exp plots 4
#| fig-cap: Proportion of Obesity across Education Levls

gridExtra::grid.arrange(plot.edu, ncol = 1)
```


From this we can see that the proportion of obesity does seem to vary across education levels, with the higher that level, the lower the proportion of obesity appears to be.

# Formal Analysis {#sec-FA}

## Prevalance of Obesity Through the Years





The logistic regression model is given by:
```{=tex}
\begin{align}
y_i &\sim  \mathrm{Bin}(n_i,p_i) \\
\end{align}
```
```{=tex}
\begin{align}
logit(p_i) = \alpha + \beta_{year2009} \cdot \mathbb{I}_{\mathrm{year2009}} + \beta_{year2010} \cdot \mathbb{I}_{\mathrm{year2010}} + \beta_{year2011} \cdot \mathbb{I}_{\mathrm{year2011}} + \beta_{year2012} \cdot \mathbb{I}_{\mathrm{year2012}}
\end{align}
```
where $\mathbb{I}_{\mathrm{year}}(x)$ is an indicator function such that
$$
\mathbb{I}_{\mathrm{year}}(x) = 
\begin{cases}
1 ~ \text{if the xth observation belongs to the particular year}, \\
0 ~ \text{otherwise}
\end{cases}
$$
Where:\
- $\text logit(p_i)$ is the log-odds of the probability of being classified as obese.\
- $\alpha$ is the intercept (baseline log-odds when predictors are zero, i.e., the Year 2008).\
- $\beta_1, \beta_2, \beta_3, \beta_4$ are the coefficients for the indicator variables corresponding to Years 2009, 2010, 2011, and 2012, respectively.\

```{r}
#| message: false
#| warning: false
data$Obese <- as.factor(data$Obese)
model <- glm(Obese ~ Year, data = data, family = binomial)

# Create a styled kable table
model %>%
  tidy(conf.int = T) %>%
  kable(format = "html", digits = 3, caption = "Logistic Regression Summary") %>%
  kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed")
  ) 
#%>%
#  column_spec(2, bold = TRUE, color = "blue") %>%
#  column_spec(3:5, color = "red")


plot_model(model, show.values = TRUE,
           title = "Odds (being obese compared to Year 2008)", show.p = FALSE)
```
The fitted model is:
$$
logit(p_i) = -0.867901  -0.037769 \cdot \mathbb{I}_{\mathrm{year2009}} + 0.043062 \cdot \mathbb{I}_{\mathrm{year2010}} + 0.003814 \cdot \mathbb{I}_{\mathrm{year2011}} + 0.001743 \cdot \mathbb{I}_{\mathrm{year2012}}$$

```
The fitted model is:
$$
logit(p_i) = -0.867901  -0.037769 \cdot \mathbb{I}_{\mathrm{year2009}} + 0.043062 \cdot \mathbb{I}_{\mathrm{year2010}} + 0.003814 \cdot \mathbb{I}_{\mathrm{year2011}} + 0.001743 \cdot \mathbb{I}_{\mathrm{year2012}}
$$


```{r}
#| message: false
#| warning: false

plot_model(model, show.values = TRUE,
           title = "Odds (being obese compared to Year 2008)", show.p = FALSE)
```
$$

- We have fitted a logistic regression model to observe the relation between the years and obesity prevalance in Scotland over the years 2008-2012. \
- The baseline category is taken to be Year 2008. On exponentiating the logistic model summary, we can see that the odds of being obese in Year 2008 was 0.42.\
- The coefficients for Year2009, Year2010, Year2011, Year2012 represents the change in odds of being obese compared to the baseline year 2008.\
- For Year 2009, the odds ratio of obesity compared to the baseline year (2008) is 0.963. This would indicate that the odds of being classified as obese in 2009 were 3.7% lower than in 2008. \
- For Year 2010, the odds ratio of obesity compared to the baseline year (2008) is 1.044. This indicates that the odds of being classified as obese in 2010 were 4.4% higher than in 2008. In other words, individuals in the year 2010 had a slightly increased likelihood of being obese compared to those in the baseline year. \

## Differences in Obesity Through The Years



# Conclusions {#sec-con}
